Reciprocal Tolerance: Why States Allow Their Adversaries to Practice Espionage with Minimal Consequences

### 2021-04-17 article.tex
Rewrite introduction

\maketitle
\newpage
\input{abstract.tex}
\newpage
\section{Introduction}
In 2015, federal authorities discovered a massive data breach inside the United States Office of Personnel Management (OPM). Hackers had, over the course of months, downloaded reams of sensitive data, including a cache of applications for federal security clearance. These applications exposed not just the personal information of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses---compromising the information of 22.1 million people in total.\footcite{nakashima_hacks_2015} US officials feared that the stolen data could help foreign intelligence services unmask undercover CIA officers, and the information could identify American officials who might be susceptible to blackmail, or worse, recruitment. \enquote{[I]f you have my [application for security clearance],} FBI Director James Comey admitted, ``you know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States. Just imagine if you were a foreign intelligence service and you had that data.''\footcite{nakashima_hacks_2015} In the case of the OPM hack, a vivid imagination wasn't necessary; the US had already established that the Chinese military was likely behind the sophisticated attack.\footcite{koerner_inside_2016} But even though it possessed conclusive digital forensics, the Obama administration refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018. See \cite{sanger_trump_2018}}

The US government's tepid response to the OPM hack illustrates a potentially surprising aspect of \nth{21} century diplomacy: when a cyberattack looks like a traditional intelligence operation, little to no diplomatic consequences will follow. Within political and academic circles, this phenomenon is often taken for granted; cyber-espionage is treated like \enquote{traditional} espionage because they are both forms of espionage. The scale of the OPM hack, however, stretches our understanding of the kind of damage that espionage can achieve. ``This is one of those cases where you have to ask, `Does the size of the operation change the nature of it?'\thinspace'' said one anonymous senior intelligence official. \enquote{Clearly, it does.}\footcite{sanger_u.s._2015} The actions of the US government say otherwise: at least for now, even operations on the scale of China's OPM infiltration still do not merit significant retaliation.

To evaluate whether espionage norms should apply in cyberspace, one first needs to understand espionage—--specifically, why espionage has traditionally been met with few diplomatic consequences. Though not codified in US policy, the light touch applied to espionage is an open secret. A senior White House official told the Congressional Research Service that \enquote{there is a vast distinction between intelligence-gathering activities that all countries do and [other forms of malicious cyber-activity].}\footnote{In the context of the quote, the White House official is distinguishing between traditional espionage and economic espionage, the latter of which is the state-sponsored theft of trade secrets from foreign private companies in order to gain a competitive advantage in international markets. See \cite{finklea_cyber_2015}} For activities that fall under the umbrella of intelligence-gathering, counterintelligence rather than prosecution is considered the appropriate response. CIA Director John Brennan once described foreign spying on domestic political institutions as \enquote{fair game.}\footcite{sanger_u.s._2016} And while speaking at an intelligence conference, James Clapper, Director of National Intelligence, was particularly frank: \enquote{You have to kind of salute the Chinese for what they did [to OPM]. If we had the opportunity to do that, I don’t think we’d hesitate for a minute.}\footcite{pepitone_clapper_2015} This permissive attitude is norm that governs espionage diplomacy, and its strategic purpose is rarely questioned or justified.

@@

This article posits that the universal tolerance of espionage in the international system can be explained through a rational calculation of each state's incentives. On balance, all states derive benefits from preserving a system in which the consequences for espionage are minimized, so they refrain from imposing particularly harsh punishments for its practice. Even when an individual incident is especially damaging to a state's security, if it falls within the traditionally accepted boundaries of an espionage operation, it will likely go unpunished. In the case of the OPM hack, for example, US policymakers decided they could not impose harsh penalties on the Chinese because it was more valuable for the US to retain its ability to practice cyber-espionage with relative impunity in the future. This basic calculation---that espionage against one's own country is a tolerable price to pay for being allowed to practice espionage against one's rivals---has remained consistent since the advent of the modern peacetime intelligence service, perhaps even the nation-state itself.

To evaluate each state's incentive, I first establish a theoretical framework for analyzing the role that espionage plays in international relations (IR) and the diplomatic norms that are associated with it. Through this framework, I define the bounds of traditional espionage and why rational states might calculate that it is in their best interests to continue tolerating it. I then examine two cases in which new technical means of gathering intelligence were introduced during the Cold War: aerial photography (spy planes) and satellite photography (spy satellites). In both cases, the norms of traditional espionage ultimately prevailed, minimizing the diplomatic consequences for their use as long as it was in service of traditional intelligence goals. Through process tracing, I demonstrate how the justifications policymakers employed for the use of these intelligence technologies---and the resulting minimization of the diplomatic penalties associated with them---are consistent with the theoretical underpinnings for a mutually strategic tacit accord permitting peacetime espionage.

@@

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Sat Aug 24 13:46:46 2019
\begin{table}[ht]
\centering
\begin{tabular}{lrrr}
  \hline
Type & Incidents & Official responses & Response \% \\
  \hline
Defacement &   3 &   2 & 66.67 \\
  Data destruction &   8 &   5 & 62.50 \\
  Doxing &   6 &   3 & 50.00 \\
  Sabotage &  14 &   5 & 35.71 \\
  Denial of service &  17 &   6 & 35.29 \\
  Espionage & 231 &  44 & 19.05 \\
   \hline
\end{tabular}
\caption{Global cyber-incident response rates}
\label{response-pct}
\end{table}

While most of the information that we have on espionage norms is qualitative, there is some data to support the limited consequences of espionage norms as well. Once again we can use cyber-espionage as an analogue for traditional espionage. Of the hundreds of state-sponsored cyberattacks known to the public, attacks classified as espionage are by far the most common. Cyber-espionage is, however, easily the least likely form of cyberattack to garner an official response from the attacked state (Table \ref{response-pct}).\footnote{These numbers were compiled from the Council on Foreign Relations' Cyber Operations Tracker, which is a publicly-sourced record of state-sponsored cyberattacks, updated quarterly. Of the 323 listed incidents, 38 were excluded from this calculation because they lacked an identifiable Victim, Sponsor, or Type. For more, see \cite{council_on_foreign_relations_new_2019}} In the world of cyber-incidents, if it looks like espionage then it is espionage, and it will receive the customary diplomatic benefits associated with that characterization.\footnote{In programming, this is a principle called \enquote{Duck Typing.} If it looks like a duck, and it quacks like a duck, then it is a duck.}

>>>>

\maketitle
\newpage
\input{abstract.tex}
\newpage
\section{Introduction}
In 2015, federal authorities discovered a massive data breach inside the United States Office of Personnel Management (OPM). Hackers had, over the course of months, downloaded reams of sensitive data, including a cache of applications for federal security clearance. These applications exposed not just the personal information of the applicants themselves but also the detailed information they had supplied about their family and friends, including social security numbers, job applications, and home addresses---compromising the information of 22.1 million people in total.\footcite{nakashima_hacks_2015} US officials feared that the stolen data could help foreign intelligence services unmask undercover CIA officers, and the information could identify American officials who might be susceptible to blackmail, or worse, recruitment. FBI Director James Comey admitted that, with the kind of information leaked, a foreign intelligence service could ``know every place I've lived since I was 18, contact people at those addresses, neighbors at those addresses, all of my family, every place I've traveled outside the United States.''\footcite[Comey is referring specifically to his SF-86, the form that one files with OPM to apply for a security clearance.]{nakashima_hacks_2015} In the case of the OPM hack, it wasn't necessary to imagine who was in possession of that information; the US had already established that the Chinese military was likely behind the sophisticated attack.\footcite{koerner_inside_2016} But even though it possessed conclusive digital forensics, the Obama administration refused to name China as the culprit.\footnote{National Security Advisor John Bolton was the first American official to formally acknowledge that the Chinese were behind the OPM hack, three years later, in September 2018. See \cite{sanger_trump_2018}}

The OPM hack was not the first time the United States Federal Government suffered a data breach at the hands of a foreign adversary, nor would it be last. Cases like the OPM hack, where sensitive data is stolen but no physical damage is done, are typically classified as \enquote{espionage;} according to the Council on Foreign Relations, there have been 73 such state-sponsored cyberattacks against the United States from 2010 through 2020 (Table \ref{response-pct}).\footnote{These numbers were compiled from the Council on Foreign Relations' Cyber Operations Tracker, which is a publicly-sourced record of state-sponsored cyberattacks, updated quarterly. The data chosen for this analysis spans from the beginning of 2010 through the end of 2020. For more, see \cite{council_on_foreign_relations_new_2019}} Of those 73 instance of cyber-espionage, only six received any public response at all. By contrast, both of the state-sponsored cyberattacks classified as \enquote{sabotage} received official responses from the US government.

Espionage is by far the most common form of cyberattack and governments do not always reveal when they have been compromised, so it is not terribly convincing to say that most cyber-espionage was ignored by the US Government while all 2 out 2 incidents of cyber-sabotage were not. But if we expand our scope globally for just a moment, it becomes apparent that the same pattern broadly holds. Stripped of all other context, espionage is the single least likely form of cyberattack to generate any type of public response.

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Sat Apr  3 15:30:42 2021
\begin{table}[ht]
\centering
\setlength{\extrarowheight}{2pt}
\small
\begin{tabular}{lrrr}
  \hline
Type & Incidents & Public responses & Response \% \\
  \hline
Defacement &   5 &   3 & 60.00 \\
  Doxing &   6 &   3 & 50.00 \\
  Data destruction &  11 &   5 & 45.45 \\
  Sabotage &  22 &   9 & 40.91 \\
  Financial Theft &   8 &   2 & 25.00 \\
  Denial of service &  13 &   3 & 23.08 \\
  Espionage & 332 &  58 & 17.47 \\
   \hline
\end{tabular}
\caption{Global cyber-incident response rates, 2010-2020}
\label{response-pct}
\end{table}

Based on this data, one could hypothesize that espionage does not always merit the same level of retaliation that sabotage or financial theft do because espionage is the least damaging form of cyberattack. While this makes sense in theory, it is not obviously true of the cyber-sabotage we've seen in practice. Consider how the OPM hack compares to a contemporary incident of state-sponsored sabotage. In 2014, hackers affiliated with the North Korean government leaked damaging emails from Sony executives and wiped some of Sony's servers.\footcite[This hack was allegedly in retaliation for the upcoming James Franco/Seth Rogen comedy about assassinating North Korean leader Kim Jong-Un.]{barnes_sony_2014} White House Press Secretary Josh Earnest alleged that the Sony hack was \enquote{destructive activity with malicious intent that was initiated by a sophisticated actor,} which he said merits \enquote{a proportional response.}\footcite{nakashima_white_2014} The Sony hack would be the first time a sitting US president publicly accused a foreign nation of launching a cyberattack and promised retaliation.\footcite{sanger_u.s._2016} The US made good on its threat and imposed sanctions on North Korea the following year.\footcite{lederman_us_2015} So why did the Sony hack---a newsworthy but far less widespread data breach---receive a significantly more serious response than a direct attack on a federal office that stole the private information of 22 million Americans?

The key distinction between the two events is the way that they are characterized by policymakers. Characterizations are not necessarily a first principle---one could choose which response is most politically expedient, and then characterize the incident accordingly---but in this case the two incidents are different in ways that would not work as an \emph{ex post facto} justification. President Obama described the Sony hack as \enquote{cyber vandalism} and Republican senators described it as \enquote{terrorism.}\footcite{fung_obama_2014} The OPM hack was instead described by the administration as a form of \enquote{intelligence-gathering activities that all countries do.}\footcite{finklea_cyber_2015} John Brennan, Director of the CIA, described foreign spying on domestic political institutions as ``fair game.''\footcite{sanger_u.s._2016} Displaying a candor atypical of the intelligence community,  Director of National Intelligence James Clapper ruefully admitted: ``[Y]ou have to kind of salute the Chinese for what they did. If we had the opportunity to do that, I don't think we'd hesitate for a minute.''\footcite{pepitone_clapper_2015}

To the US government, stealing classified data from a federal office looks like something a spy might do, but wiping a movie studio's servers does not. While this explanation might ring true---cyber-espionage is espionage, sure---it is also essentially a tautology. It raises the question: why does it matter whether something qualifies as espionage, and is that a label that continues to apply even when the scale of what espionage can accomplish is magnified by new technology? Like a traditional intelligence operation, the OPM hack \enquote{merely} involved stealing sensitive information, but the quantity of compromised data was orders of magnitude beyond what could be accomplished by photographing the contents of a file cabinet and passing it along in a briefcase. \enquote{This is one of those cases where you have to ask, `Does the size of the operation change the nature of it?'\thinspace} said one anonymous senior intelligence official. \enquote{Clearly, it does.}\footcite{sanger_u.s._2015} The actions of the US government still say otherwise: at least for now, even operations on the scale of China's OPM infiltration do not merit significant retaliation.

This article seeks to explain how an old set of espionage norms continues to weather disruptive technological change. To do so first requires one to explain how it is possible that the routine minimization of espionage response happens at all. Extant espionage literature tends to focus on prescriptive questions---should espionage exist? is espionage legal?---rather than descriptive ones. There are few academic scholars who examine the theoretical implications of espionage in practice: what does it mean for existing theories of IR that a category of inter-state action routinely receives retaliation far less severe than the action's national security implications would seem to merit? In the first section of this paper I review the existing scholarship on norms and note the areas in which espionage does not neatly fit into the existing research.

In the latter section of this paper I demonstrate how espionage norms have been deployed during previous periods of tectonic technological change. The historical record demonstrates that the idea of espionage is deeply meaningful to the individual actors that decide state policy; if an action is considered to be espionage, it exists in a self-contained ontological space, with rules and consequences unaffected by rational cost analysis. The people who are in charge of these matters---what operations to undertake, how much risk to absorb, whether to respond to an operation conducted against them---intuit whether something \emph{counts as} espionage, and adjust their own actions to fit within that paradigm. The benefits for doing so are clear: if the aggrieved party considers an operation to be espionage, it is much less likely to retaliate, even when the consequences for that operation are deeply damaging. Cyber-espionage is a novel challenge for the intelligence community, but it is not the first paradigmatic shift in intelligence capabilities. Spy planes and spy satellites radically re-defined what espionage could accomplish and changed the course of the Cold War---the consequences for using them did not evolve accordingly.

\newpage

@@

% This article posits that the universal tolerance of espionage in the international system can be explained through a rational calculation of each state's incentives. On balance, all states derive benefits from preserving a system in which the consequences for espionage are minimized, so they refrain from imposing particularly harsh punishments for its practice. Even when an individual incident is especially damaging to a state's security, if it falls within the traditionally accepted boundaries of an espionage operation, it will likely go unpunished. In the case of the OPM hack, for example, US policymakers decided they could not impose harsh penalties on the Chinese because it was more valuable for the US to retain its ability to practice cyber-espionage with relative impunity in the future. This basic calculation---that espionage against one's own country is a tolerable price to pay for being allowed to practice espionage against one's rivals---has remained consistent since the advent of the modern peacetime intelligence service, perhaps even the nation-state itself.

% To evaluate each state's incentive, I first establish a theoretical framework for analyzing the role that espionage plays in international relations (IR) and the diplomatic norms that are associated with it. Through this framework, I define the bounds of traditional espionage and why rational states might calculate that it is in their best interests to continue tolerating it. I then examine two cases in which new technical means of gathering intelligence were introduced during the Cold War: aerial photography (spy planes) and satellite photography (spy satellites). In both cases, the norms of traditional espionage ultimately prevailed, minimizing the diplomatic consequences for their use as long as it was in service of traditional intelligence goals. Through process tracing, I demonstrate how the justifications policymakers employed for the use of these intelligence technologies---and the resulting minimization of the diplomatic penalties associated with them---are consistent with the theoretical underpinnings for a mutually strategic tacit accord permitting peacetime espionage.

@@

While most of the information that we have on espionage norms is qualitative, there is some data to support the limited consequences of espionage norms as well. Once again we can use cyber-espionage as an analogue for traditional espionage. Of the hundreds of state-sponsored cyberattacks known to the public, attacks classified as espionage are by far the most common. Cyber-espionage is, however, easily the least likely form of cyberattack to garner an official response from the attacked state

In the world of cyber-incidents, if it looks like espionage then it is espionage, and it will receive the customary diplomatic benefits associated with that characterization.\footnote{In programming, this is a principle called \enquote{Duck Typing.} If it looks like a duck, and it quacks like a duck, then it is a duck.}
